---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# simpleExpressionProcessing

<!-- badges: start -->
<!-- badges: end -->

The goal of simpleExpressionProcessing is to provide simple tools for downloading and normalizing raw RNA expression data from GEO, including microarray and bulk RNA sequencing experiments, within the `GEOquery` class format. 


## Installation

You can install the development version of simpleExpressionProcessing from [GitHub](https://github.com/) with:

``` r
devtools::install_github("BenSolomon/simpleExpressionProcessing")
```

## Example usage

### Stepwise approach

This workflow involves the following steps:

- GEO data is downloaded with `MetaIntegrator`. This includes an `expr` slot which contains the expression data, as normalized by the data authors
- Supplemental data files are downloaded with `GEOquery`, such as raw `.CEL` files from the Affymetrix microarray platform
- The raw expression files are normalized 
- The GEO data object is modified in the following ways:
  - The original `expr` slot is reassigned to the `exprGEO` slot, to indicate it represents the author normalized data
  - The raw, non-normalized data from the supplemental files are assigned to the `exprRAW` slot
  - The user normalized data is added to the `expr` slot, which is the slot used by downstream applications using MetaIntegrator GEO objects
- All expression slots (i.e. `expr`, `exprRAW`, and `exprGEO`) are visualized with boxplots to evaluate the normalization process

```{r}
devtools::load_all() # Only needed in package development environment
accession <- "GSE16797"

# Get GEO daat
gse <- MetaIntegrator::getGEOData(accession)

# Get GEO supplemental data, including raw data, and process
## withr used to create a tempdir for supplemental files for purpose of demo
withr::with_tempdir({
  getSuppFiles(accession)
  reprocess_output <- processAffy(sprintf("./%s", accession))
  
  # # Alternatively, can use `processAgilent` for Agilent platform
  # reprocess_output <- processAgilent(sprintf("./%s", accession)) 
  
})

# Add raw and reprocessed expression matrices to original MetaIntegrator
# GEO object
gse <- addRawExprMatrix(gse, reprocess_output$raw_expression)
gse <- addReprocessedExprMatrix(gse, reprocess_output$normalized_expression)

# Visualize effect of processing
geoAllBoxplots(gse)
```

